Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex

\textbf{Reward predict is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learning loss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\hat{z}_{t+h} &= d_{\phi}(\hat{z}_{t}, a_{t}) \\
		\tilde{z}_{t+h} &= e_{\bar{\theta}}(s_{t+1}) \\
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right )
	\end{align*}
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to being able to maintain the latent matrix rank in this specific environment (see \cref{fig:rank-comparison-z1}). Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

@@

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\hat{o}_t &= g_\xi(z_t), \\
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:multi-task-pretraining} show that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the overall performance, thus lending further support to the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning task-agnostic representations which can be combined with
any model-free RL algorithm.
In our experiments, we compared \our to both state-of-the-art model-free RL methods and state-of-the-art model-based RL methods.

Our results demonstrate state-of-the-art
based on our empirical results, we suggest further

\textbf{Limitations and future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research into multi-task and lifelong learning.
Given that our method obtains an implicit codebook, an obvious future direction is exploring whether we can leverage the
discretized latent-space. So far, we have only applied our representation loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an attractive direction. Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample-efficient. Exploring whether explicitly quantifying the model uncertainty by \eg using larger ensembles would help to adapt our algorithm to a real-world robotics setting is also a promising direction for future work.

>>>>

\textbf{Reward predict is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learning loss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right )
	\end{align*}
  where as before we have $\hat{z}_{t+h} = d_{\phi}(\hat{z}_{t}, a_{t})$ and $\tilde{z}_{t+h} = e_{\bar{\theta}}(s_{t+1})$.
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to being able to maintain the latent matrix rank in this specific environment (see \cref{fig:rank-comparison-z1}). Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

@@

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\hat{o}_t &= g_\xi(z_t), \\
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} shows that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the overall performance, thus lending further support to the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning representations which demonstrates state-of-the-art performance in
continuous control tasks.
Importantly, our normalization and bounding of the latent space empirically preserves the representation's matrix rank,
indicating that it alleviates representation collapse.
Importantly, our method is {\em (i)} straightforward, {\em (ii)} compatible with any model-free RL algorithm,
and {\em (iii)} learns a task-agnostic representation.
% Moreover, it can be combined with any model-free RL algorithm
% In our experiments we compared against both state-of-the-art model-free RL methods and state-of--the-art model-based RL methods.
% Our results demonstrate state-of-the-art based on our empirical results, we suggest further

\textbf{Limitations and future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research into multi-task and lifelong learning.
% Given that our method obtains an implicit codebook an obvious future direction is exploring if we can leverage the
% discretized latent-space. So far, we have only applied our representation learning loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an interesting direction.
% Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample efficient. Exploring whether explicitly quantifying the model uncertainty by \eg\ using larger ensembles would help to adapt our method to a real-world robotics setting is also a promising direction for future work.
Given that our method obtains an implicit codebook, an obvious future direction is exploring whether we can leverage the
discretized latent-space. So far, we have only applied our representation loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an attractive direction.
% Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample-efficient. Exploring whether explicitly quantifying the model uncertainty by \eg using larger ensembles would help to adapt our algorithm to a real-world robotics setting is also a promising direction for future work.
